---
title: "useR! 2020 Tutorial Part 1"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Intro

### What is disk.frame for?

### Installation

```{r installation, exercise = TRUE}
# to install CRAN version
install.packages("disk.frame")

# Uncomment below to install latest development version
# install.packages("remotes")
# remotes::install_github("xiaodaigh/disk.frame")
```

On some platforms, it helps to specify the repo if you encounter an 
```r
install.packages("disk.frame", repo="https://cran.rstudio.com")
```
### Download data?

### Generate data?


## Starting disk.frame

### Exercise 

{disk.frame} can use multiple cores and may transfer data between **workers**.

So there are some setting up code that is recommended for all users.

```{r setting_up, exercise=TRUE}
library(disk.frame)

```

You should see a message like this from the console
```
# this will set up disk.frame with multiple workers
setup_disk.frame()
# this will allow unlimited amount of data to be passed from worker to worker
options(future.globals.maxSize = Inf)
```


```{r setting_up1, exercise=TRUE}
# this will set up disk.frame with multiple workers
setup_disk.frame()
# this will allow unlimited amount of data to be passed from worker to worker
options(future.globals.maxSize = Inf)
```

If you forget

```{r setting_up2, exercise=TRUE}
disk.frame::show_boilerplate()

# or equivalently
disk.frame::show_ceremony()
```

TODO insert GIF of inserting code into RStudio.

## Ingesting Data

To use {disk.frame} as your data processing engine, you need to store your data in the disk.frame format.

### The `disk.frame` format

The {disk.frame} format is a folder containing multiple [`fst`](https://www.fstpackage.org/) files. And the files are named `1.fst`, `2.fst` etc, where each `fst` file is called a **chunk**.

TODO add quiz

```{r ingest_data_frame, exercise=TRUE}
mtcars.df = as.disk.frame(mtcars)
mtcars.df
```

This works but if your data fits into RAM, you probably don't need {disk.frame}. But perhaps you data.frame fits into RAM but you can't merge it with another dataset, because that would require additional RAM. So storing it as disk.frame before merging can help.

## Ingesting CSV

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
data.table::fwrite(mtcars, "mtcars.csv")
# alternatively
#write.csv(mtcars, "mtcars.csv")

mtcars.df = csv_to_disk.frame("mtcars.csv")
mtcars.df
```

Notice how `mtcars.df` are stored in a temporary folder? You can move it using

```{r}
move_to(mmtcars.df, "/some/path/permanent")
```

or 

```{r}
mtcars.df = csv_to_disk.frame(
  "mtcars.csv", 
  outdir = "/some/path/permanent")

lib1 = libname("some_folder")

lib1$mtcars2.df = lib1$mtcars.df %>% filter(mpg < 0.5)
```



### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

