---
title: "Generalized Linear Models (GLM) including logistic regression with disk.frame"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generalized Linear Models (logistic regression etc) with disk.frame}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, cache=TRUE}
suppressPackageStartupMessages(library(disk.frame))

if(interactive()) {
  setup_disk.frame() 
} else {
  # only use 1 work to pass CRAN check
  setup_disk.frame(1)
}

```

# GLMs

### Prerequisites
In this article, we will assume you are familiar with Generalized Linear Models (GLMs). You are also expected to have basic working knowledge of {`disk.frame`}, see this [{`disk.frame`} Quick Start](http://diskframe.com/articles/intro-disk-frame.html). 

## Introduction
One can fit a GLM using the `glm` function. For example,

```{r glm, cache=TRUE}
m = glm(dist ~ speed, data = cars)
```

would fit a linear model on the data `cars` with `dist` as the target and `speed` as the explanatory variable. You can inspect the results of the model fit using

```{r, depeondson='glm'}
summary(m)
```

or if you have `{broom}` installed

```{r, depeondson='glm'}
broom::tidy(m)
```

With {`disk.frame`}, you can run GLM `dfglm` function, where the `df` stands for `disk.frame` of course!
```{r dependson='setup'}
cars.df = as.disk.frame(cars)

m = dfglm(dist ~ speed, cars.df)

summary(m)


majorv = as.integer(version$major)
minorv = as.integer(strsplit(version$minor, ".", fixed=TRUE)[[1]][1])

if((majorv == 3) & (minorv >= 6)) {
  broom::tidy(m)
} else {
  # broom doesn't work in version < R3.6 because biglm does not work
}

```

The syntax didn't change at all! You are able to enjoy the benefits of `disk.frame` when dealing with larger-than-RAM data.

## Logistic regression
Logistic regression is one of the most commonly deployed machine learning (ML) models. It is often used to build binary classification models

```{r dependson='setup'}
iris.df = as.disk.frame(iris)

# fit a logistic regression model to predict Speciess == "setosa" using all variables
all_terms_except_species = setdiff(names(iris.df), "Species")
formula_rhs = paste0(all_terms_except_species, collapse = "+")

formula = as.formula(paste("Species == 'versicolor' ~ ", formula_rhs))

iris_model = dfglm(formula , data = iris.df, family=binomial())

# iris_model = dfglm(Species == "setosa" ~ , data = iris.df, family=binomial())

summary(iris_model)

majorv = as.integer(version$major)
minorv = as.integer(strsplit(version$minor, ".", fixed=TRUE)[[1]][1])

if((majorv == 3) & (minorv >= 6)) {
  broom::tidy(iris_model)
} else {
  # broom doesn't work in version < R3.6 because biglm does not work
}

```

The arguments to the `dfglm` function are the same as the arguments to `biglm::bigglm` which are based on the `glm` function. Please check their documentations for other argument options.

## Notes
`{disk.frame}` uses `{biglm}` and `{speedglm}` as the backend for GLMs. Unfortunately, neither package is managed on open-source platforms, so it's more difficult to contribute to them by making bug fixes and submitting bug reports. So bugs are likely to persists. There is an active effort on `disk.frame` to look for alternatives. Example of avenues to explore include tighter integration with `{keras}`, h2o, or Julia's OnlineStats.jl for model fit purposes.

Another package for larger-than-RAM glm fitting, `{bigFastlm}`, has been taken off CRAN, it is managed on Github.

Currently, parallel processing of GLM fit are not possible with {`disk.frame`}. 
