---
title: "Quick Start: Basic Operations with nycflights13"
author: "ZJ"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
suppressPackageStartupMessages(library(disk.frame))
library(fst)
library(magrittr)
library(nycflights13)
library(dplyr)
library(data.table)

# you need to run this for multi-worker support
# limit to 2 cores if not running interactively; most likely on CRAN
# set-up disk.frame to use multiple workers
if(interactive()) {
  setup_disk.frame()
  # highly recommended, however it is pun into interactive() for CRAN because
  # change user options are not allowed on CRAN
  options(future.globals.maxSize = Inf)  
} else {
  setup_disk.frame(2)
}


knitr::opts_chunk$set(
	eval = FALSE,
	collapse = TRUE,
	comment = "#>",
	include = TRUE
)
```

# Quick Start - replicating dplyr's tutorial on nycflight13 

The [`disk.frame` package](https://github.com/xiaodaigh/disk.frame) aims to be the answer to the question: how do I manipulate structured tabular data that doesn't fit into Random Access Memory (RAM)? 

In a nutshell, `disk.frame` makes use of two simple ideas

1) split up a larger-than-RAM dataset into chunks and store each chunk in a separate file inside a folder and 
2) provide a convenient API to manipulate these chunks

`disk.frame` performs a similar role to distributed systems such as Apache Spark, Python's Dask, and Julia's JuliaDB.jl for *medium data* which are datasets that are too large for RAM but not quite large enough to qualify as *big data*.

In this tutorial, we introduce `disk.frame`, address some common questions, and replicate the [sparklyr data manipulation tutorial](https://spark.rstudio.com/dplyr/) using `disk.frame` constructs.

## Installation
Simply run

```r
install.packages("disk.frame") # when CRAN ready
```
or

```r
devtools::install_github("xiaodaigh/disk.frame")
```

## Set-up `disk.frame`
`disk.frame` works best if it can process multiple data chunks in parallel. The best way to set-up `disk.frame` so that each CPU core runs a background worker is by using

```r
setup_disk.frame()

# this will allow unlimited amount of data to be passed from worker to worker
options(future.globals.maxSize = Inf)
```

The `setup_disk.frame()` sets up background workers equal to the number of CPU cores; please note that, by default, hyper-threaded cores are counted as one not two.

Alternatively, one may specify the number of workers using `setup_disk.frame(workers = n)`.

## Basic Data Operations with `disk.frame`

The `disk.frame` package provides convenient functions to convert `data.frame`s and CSVs to `disk.frame`s.

### Creating a `disk.frame` from `data.frame`
We convert a `data.frame` to `disk.frame` using the `as.data.frame` function.

```{r asdiskframe, cache=TRUE}
library(nycflights13)
library(dplyr)
library(disk.frame)
library(data.table)

# convert the flights data to a disk.frame and store the disk.frame in the folder
# "tmp_flights" and overwrite any content if needed
flights.df <- as.disk.frame(
  flights, 
  outdir = file.path(tempdir(), "tmp_flights.df"),
  overwrite = TRUE)
flights.df
```
You should now see a folder called `tmp_flights` with some files in it, namely `1.fst`, `2.fst`.... where each `fst` files is one chunk of the `disk.frame`.


### Creating a `disk.frame` from CSV
```{r}
library(nycflights13)
# write a csv
csv_path = file.path(tempdir(), "tmp_flights.csv")
data.table::fwrite(flights, csv_path)

# load the csv into a disk.frame
df_path = file.path(tempdir(), "tmp_flights.df")
flights.df <- csv_to_disk.frame(
  csv_path, 
  outdir = df_path,
  overwrite = T)
  
flights.df
```

If the CSV is too large to read in, then we can also use the `in_chunk_size` option to control how many rows to read in at once. For example to read in the data 100,000 rows at a time.

```{r}
library(nycflights13)
library(disk.frame)

# write a csv
csv_path = file.path(tempdir(), "tmp_flights.csv")

data.table::fwrite(flights, csv_path)

df_path = file.path(tempdir(), "tmp_flights.df")

flights.df <- csv_to_disk.frame(
  csv_path, 
  outdir = df_path, 
  in_chunk_size = 100000)
  
flights.df
```

`disk.frame` also has a function `zip_to_disk.frame` that can convert every CSV in a zip file to `disk.frame`s.

### Simple `dplyr` verbs and lazy evaluation
```{r dfselect, dependson='asdiskframe', cache=TRUE}
flights.df1 <- select(flights.df, year:day, arr_delay, dep_delay)
flights.df1
```

```{r dependson='dfselect'}
class(flights.df1)
```

The class of `flights.df1` is also a `disk.frame` after the   `dplyr::select` transformation. Also, `disk.frame` operations are by default (and where possible) **lazy**, meaning it doesn't perform the operations right away. Instead, it waits until you call `collect`. Exceptions to this rule are the `*_join` operations which evaluated *eagerly* under certain conditions see **Joins for disk.frame in-depth** for details.

For lazily constructed `disk.frame`s (e.g. `flights.df1`). The function `collect` can be used to bring the results from disk into R, e.g.
```{r, dependson='dfselect'}
collect(flights.df1) %>% head(2)
```

Of course, for larger-than-RAM datasets, one wouldn't call `collect` on the whole `disk.frame` (because why would you need `disk.frame` otherwise). More likely, one would call `collect` on a `filter`ed dataset or one summarized with `group_by`.

Some examples of other dplyr verbs applied:

```{r, dependson='asdiskframe'}
filter(flights.df, dep_delay > 1000) %>% collect %>% head(2)
```

```{r, dependson='asdiskframe'}
mutate(flights.df, speed = distance / air_time * 60) %>% collect %>% head(2)
```

### Examples of NOT fully supported `dplyr` verbs

The `chunk_arrange` function arranges (sorts) each chunk but not the whole dataset. So use with caution. Similarly `chunk_summarise` creates summary variables within each chunk and hence also needs to be used with caution. In the Group By section, we demonstrate how to use `summarise` in the `disk.frame` context correctly with `hard_group_by`s.

```{r, dependson='asdiskframe'}
# this only sorts within each chunk
chunk_arrange(flights.df, dplyr::desc(dep_delay)) %>% collect %>% head(2)
```


```{r, dependson='asdiskframe'}
chunk_summarize(flights.df, mean_dep_delay = mean(dep_delay, na.rm =T)) %>% collect
```

### Piping

One can chain `dplyr` verbs together like with a `data.frame`

```{r, dependson='asdiskframe'}
c4 <- flights %>%
  filter(month == 5, day == 17, carrier %in% c('UA', 'WN', 'AA', 'DL')) %>%
  select(carrier, dep_delay, air_time, distance) %>%
  mutate(air_time_hours = air_time / 60) %>%
  collect %>%
  arrange(carrier)# arrange should occur after `collect`

c4  %>% head
```

### List of supported `dplyr` verbs

```r
select
rename
filter
chunk_arrange # within each chunk
chunk_group_by # within each chunk
chunk_summarize # within each chunk
group_by # limited functions
summarize # limited functions
mutate
transmute
left_join
inner_join
full_join # careful. Performance!
semi_join
anit_join
```

## Sharding and distribution of chunks

Like other distributed data manipulation frameworks `disk.frame` utilizes the *sharding* concept to distribute the data into chunks. For example "to shard by `cust_id`" means that all rows with the same `cust_id` will be stored in the same chunk. This enables `chunk_group_by` by `cust_id` to produce the same results as non-chunked data.

The `by` variables that were used to shard the dataset are called the `shardkey`s. The *sharding* is performed by computing a deterministic hash on the shard keys (the `by` variables) for each row. The hash function produces an integer between `1` and `n`, where `n` is the number of chunks. 

## Group-by

`{disk.frame}` implements the `group_by` operation some caveats. In the `{disk.frame}` framework, only a set functions are supported in `summarize`. However, the user can create more custom `group-by` functions can be defined.

```{r, dependson='asdiskframe'}
flights.df %>%
  group_by(carrier) %>% # notice that hard_group_by needs to be set
  summarize(count = n(), mean_dep_delay = mean(dep_delay, na.rm=T)) %>%  # mean follows normal R rules
  collect %>% 
  arrange(carrier)
```

## Restrict input columns for faster processing

One can restrict which input columns to load into memory for each chunk; this can significantly increase the speed of data processing. To restrict the input columns, use the `srckeep` function which only accepts column names as a string vector.

```{r, dependson='asdiskframe'}
flights.df %>%
  srckeep(c("carrier","dep_delay")) %>%
  group_by(carrier) %>% 
  summarize(count = n(), mean_dep_delay = mean(dep_delay, na.rm=T)) %>%  # mean follows normal R rules
  collect
```

Input column restriction is one of the most critical efficiencies provided by `disk.frame`. Because the underlying format allows random access to columns (i.e. retrieve only the columns used for processing), hence one can drastically reduce the amount of data loaded into RAM for processing by keeping only those columns that are directly used to produce the results.

## Joins

`disk.frame` supports many dplyr joins including:

```r
left_join
inner_join
semi_join
inner_join
full_join # requires hard_group_by on both left and right
```
In all cases, the left dataset (`x`) must be a `disk.frame`, and the right dataset (`y`) can be either a `disk.frame` or a `data.frame`. If the right dataset is a `disk.frame` and the `shardkey`s are different between the two `disk.frame`s then two expensive `hard` `group_by` operations are performed *eagerly*, one on the left `disk.frame` and one on the right `disk.frame` to perform the joins correctly. 

However, if the right dataset is a `data.frame` then `hard_group_by`s are only performed in the case of `full_join`.

Note `disk.frame` does not support `right_join` the user should use `left_join` instead.

The below joins are performed *lazily* because `airlines.dt` is a `data.table` not a `disk.frame`:

```{r airlines_dt, dependson='asdiskframe', cache=TRUE}
# make airlines a data.table
airlines.dt <- data.table(airlines)
# flights %>% left_join(airlines, by = "carrier") #
flights.df %>% 
  left_join(airlines.dt, by ="carrier") %>% 
  collect %>% 
  head
```

```{r, dependson='airlines_dt'}
flights.df %>% 
  left_join(airlines.dt, by = c("carrier", "carrier")) %>% 
  collect %>% 
  tail
```

## Window functions and arbitrary functions

`{disk.frame}` supports all `data.frame` operations, unlike Spark which can only perform those operations that Spark has implemented. Hence windowing functions like `min_rank` and `rank` are supported out of the box. 

For the following example, we will use the `hard_group_by` which performs a group-by and also reorganises the chunks so that all records with the same `year`, `month`, and `day` end up in the same chunk. This is typically not advised, as `hard_group_by` can be slow for large datasets.

```{r, dependson='asdiskframe'}
# Find the most and least delayed flight each day
bestworst <- flights.df %>%
   srckeep(c("year","month","day", "dep_delay")) %>%
   hard_group_by(c("year", "month", "day")) %>%
   filter(dep_delay == min(dep_delay, na.rm = T) || dep_delay == max(dep_delay, na.rm = T)) %>%
   collect
   
bestworst %>% head
```

another example

```{r, dependson='asdiskframe'}
ranked <- flights.df %>%
  srckeep(c("year","month","day", "dep_delay")) %>%
  hard_group_by(c("year", "month", "day")) %>%
  filter(min_rank(desc(dep_delay)) <= 2 & dep_delay > 0) %>%
  collect

ranked %>% head
```

one more example

```{r, dependson='asdiskframe'}
# Rank each flight within a daily
ranked <- flights.df %>%
  srckeep(c("year","month","day", "dep_delay")) %>%
  chunk_group_by(year, month, day) %>%
  select(dep_delay) %>%
  mutate(rank = rank(desc(dep_delay))) %>%
  collect

ranked %>% head
```


## Arbitrary by-chunk processing

One can apply arbitrary transformations to each chunk of the `disk.frame` by using the `delayed` function which evaluates lazily or the `map.disk.frame(lazy = F)` function which evaluates eagerly. For example to return the number of rows in each chunk

```{r, dependson='asdiskframe'}
flights.df1 <- delayed(flights.df, ~nrow(.x))
collect_list(flights.df1) %>% head # returns number of rows for each data.frame in a list
```
and to do the same with `map.disk.frame`

```{r, dependson='asdiskframe'}
map(flights.df, ~nrow(.x), lazy = F) %>% head
```
The `map` function can also output the results to another disk.frame folder, e.g.

```{r, dependson='asdiskframe'}
# return the first 10 rows of each chunk
flights.df2 <- map(flights.df, ~.x[1:10,], lazy = F, outdir = file.path(tempdir(), "tmp2"), overwrite = T)

flights.df2 %>% head
```

Notice `{disk.frame}` supports the `purrr` syntax for defining a function using `~`.

## Sampling

In the `disk.frame` framework, sampling a proportion of rows within each chunk can be performed using `sample_frac`. 

```{r, dependson='asdiskframe'}
flights.df %>% sample_frac(0.01) %>% collect %>% head
```

## Writing Data

One can output a `disk.frame` by using the `write_disk.frame` function. E.g.

```r
write_disk.frame(flights.df, outdir="out")
```
this will output a disk.frame to the folder "out"

```{r cleanup}
fs::dir_delete(file.path(tempdir(), "tmp_flights.df"))
fs::dir_delete(file.path(tempdir(), "tmp2"))
fs::file_delete(file.path(tempdir(), "tmp_flights.csv"))
```
