---
title: "Simple Fannie Mae Example"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r}
# 2_exploratory.r
#source("../fannie_mae/00_setup.r")
source("00_setup.r")
```

# Assign a `disk.frame`
You use `disk.frame(path)` to assign a disk.frame

```{r}
dir(file.path(outpath, "fm_with_harp")) %>% 
  head
```

Use the function `disk.frame` to assign a folder
```{r}
fm_with_harp = 
  disk.frame(file.path(outpath, "fm_with_harp"))

fm_with_harp
```

## Simple info about disk.frame
```{r}
nrow(fm_with_harp)

ncol(fm_with_harp)
```


```{r}
head(fm_with_harp)
```

```{r}
tail(fm_with_harp)
```

## Simple group-by example

I want create the `sum(default_12m)` group by month. The sql statement looks like

```sql
select
  monthly.rpt.prd,
  sum(default_12m) as n_defaults,
  count(*) as cnt,
  n_default/cnt as odr # observed default rate
from
  table
group by
  monthly.rpt.prd;
```
This analysis only uses two columns, namely `default_12m` and `monthly.rpt.prd`. So I use `srckeep` to ensure that only those two columns are loaded.

The backend for disk.frame is fst which allows **random access** to columns AKA load only what you need! This is importance for performance!
```{r}
system.time(a_wh1 <- fm_with_harp %>% 
  srckeep(c("default_12m","monthly.rpt.prd")) %>% 
  group_by(monthly.rpt.prd) %>% 
  summarise(
    N = n(), 
    n_defaults = sum(default_12m, na.rm = T)))
```
The error recognises that the `shardkey` is `loan_id` which is not the group-by variable - `monthly.rpt.prd`.

The operations returned almost instantly, this is because the computation was done lazily. For now operations are recorded. No computation will take place until `collect` or `compute` is called.

Now **play** the recording by calling `collect`. This will take a minute. Back to the slides:

```{r}
system.time(a_wh2 <- a_wh1 %>% collect) # 60~70 plugged in

a_wh2
```

```{r}
# temporary
# a_wh1 %>% 
#   srckeep(c("monthly.rpt.prd", "default_12m")) %>% 
#   map(function(chunk) {
#     chunk[1,]
#   }) %>% 
#   collect
```


```{r}
# temporary
# a_wh1 %>% 
#   srckeep(c("monthly.rpt.prd", "default_12m")) %>% 
#   map(~{
#     .x[1,.SD]
#   }) %>% 
  # collect
```


once `collect` is called the resultant data is stored as a data.frame.

However this is not the correct result, as the group by was performed within each chunk. Hence we need a second stage group by. The second group by takes no time at all, as everything was done in memory

```{r}
system.time(a_wh3 <- a_wh2 %>%
  group_by(monthly.rpt.prd) %>% 
  summarise(
    odr = sum(n_defaults)/sum(N)
  ) %>% 
  rename(
    Date = monthly.rpt.prd,
    `Observed Default Rate%` = odr
  ))
```

Manipulate the data and plot
```{r}
a_wh4 = a_wh3 %>% 
  gather(key = type, value=rate, -Date)

ggplot(a_wh4) + 
  geom_line(aes(x=Date, y = rate, colour = type)) +
  ggtitle("Fannie Mae Observed Default Rate over time & HARP Conversion Rate")
```
