---
title: "Key `{disk.frame}` concepts"
author: "ZJ"
output: pdf_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Key `{disk.frame}` concepts
There are a number of concepts and terminologies that are useful to understand in order to use `disk.frame` effectively.

## What is a `disk.frame` and what are chunks?

A `disk.frame` is a folder containing [`fst`](https://www.fstpackage.org/) files named "1.fst", "2.fst", "3.fst" etc. Each of the ".fst" file is called a _chunk_.

## Workers and parallelism

Parallelism in `{disk.frame`} is achieved using the [`{future}` package](https://cran.r-project.org/package=future). When performing many tasks, `{disk.frame}` uses multiple workers to perform the tasks in parallel. A _worker_ is an R session.

It is recommended that you run the following immediately after `library(disk.frame)` to set-up multiple workers. For example:

```{r}
library(disk.frame)
setup_disk.frame()

# this will allow unlimited amount of data to be passed from worker to worker
options(future.globals.maxSize = Inf)
```

For example, suppose we wish to compute the number of rows for each chunk, we can clearly perform this simultaneously in parallel. The code to do that is

```r
# use only one column is fastest
df[,.N, keep = "first_col"]
```

or equivalent using the `srckeep` function

```r
# use only one column is fastest
srckeep(df, "first_col")[,.N, keep = "first_col"]
```

Say there are `n` chunks in `df`, and there are `m` workers. Then the first `m` chunks will run `chunk[,.N]` simultaneously.

To see how many workers are available for `{disk.frame}` to use, run
```r
future::nbrOfWorkers()
```

## How `{disk.frame}` works

When `df %>% some_fn %>% collect` is called. The `some_fn` is applied to each chunk of `df`. The collect will row-bind the results from `some_fn(chunk)`together if the returned value of `some_fn` is a data.frame, or it will return a `list` containing the results of `some_fn`.

The session that receives these results is called the **main session**. In general, we should try to minimize the amount of data passed from the worker sessions back to the main session, because passing data around can be slow.

Also, please note that there is no communication between the workers, except for workers passing data back to the main session.


