---
title: "Using data.table syntax with disk.frame"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using data.table syntax with disk.frame}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## `disk.frame` supports `data.table` syntax


```{r setup_data_table, cache=TRUE}
library(disk.frame)

# set-up disk.frame to use multiple workers
if(interactive()) {
  setup_disk.frame()
  # highly recommended, however it is pun into interactive() for CRAN because
  # change user options are not allowed on CRAN
  options(future.globals.maxSize = Inf)  
} else {
  setup_disk.frame(2)
}


library(nycflights13)

# create a disk.frame
flights.df = as.disk.frame(nycflights13::flights, outdir = file.path(tempdir(),"flights13"), overwrite = TRUE)
```

In the following example, I will use the `.N` from the `data.table` package to count the unique combinations `year` and `month` within each chunk.

```{r ok, dependson='setup_data_table'}
library(data.table)
library(disk.frame)

flights.df = disk.frame(file.path(tempdir(),"flights13"))

names(flights.df)

flights.df[,.N, .(year, month), keep = c("year", "month")]
```

All `data.table` syntax are supported. However, `disk.frame` adds the ability to load only those columns required for the analysis using the `keep =` option. In the above analysis, only the `year` and `month` variables are required and hence `keep = c("year", "month")` was used. 

Alternatively, we can use the `srckeep` function to achieve the same, e.g.

```r
srckeep(flights.df, c("year", "month"))[,.N, .(year, month)]
```

### External variables are captured

`disk.frame` sends the computaion to background workers which are essentially distinct and separate R sessions. Typically, the variables that you have available in your current R session aren't visible in the other R sessions, but `disk.frame` uses the `future` package's variable detection abilities to figure out which variables are in use and then send them to the background workers so they have access to the variables as well. E.g.

```{r var_detect, dependson='setup_data_table'}
y = 42 
some_fn <- function(x) x


flights.df[,some_fn(y)]
```

In the above example, neither `some_fn` nor `y` are defined in the background workers' environments, but `disk.frame` still manages to evaluate this code `flights.df[,some_fn(y)]`.

```{r clean_up, include=FALSE}
fs::dir_delete(file.path(tempdir(),"flights13"))
```